---
title: "Credit Card Fraud Detection"
author: "Akia Canada"
date: "2025-04-05"
output: html_document
---


Package Install

```{r}
install.packages(c("data.table", "dplyr", "ggplot2", "corrplot", "caret", "ROCR", 
                   "rpart", "nnet", "xgboost", "DMwR", "e1071", "pROC"))
install.packages("DMwR2")

```

```{r}
install.packages("DMwR2")
```



```{r}
# Load necessary libraries

library(tidyr)
library(treemapify)  # For Tree Map
library(scales)      # For percentage scales
library(stringr)  # For wrapping text

library(data.table)
library(dplyr)
library(ggplot2)
library(corrplot)
library(caret)
library(ROCR)
library(rpart)
library(nnet)
library(xgboost)
library(DMwR2)
library(e1071)
library(pROC)


```





```{r}
# Load datasets 

df_2013 <- fread("creditcard.csv")          
df_2023 <- fread("creditcard_2023.csv")      
```


```{r}
# Inspect structure

str(df_2013)
summary(df_2013)
table(df_2013$Class)
```



```{r}
# Inspect structure

str(df_2023)
summary(df_2023)
table(df_2023$Class)

```


# ========== CLEANING ==========

```{r}
# 1. Remove 'Time' column from 2013 dataset
df_2013 <- df_2013 %>% select(-Time)
```


```{r}
# 2. Scale 'Amount'
df_2013$Amount <- scale(df_2013$Amount)
```

```{r}
# 3. Check and remove NA values (if any)
df_2013 <- na.omit(df_2013)
df_2023 <- na.omit(df_2023)
```

```{r}
# 4. Convert Class to factor
df_2013$Class <- factor(df_2013$Class, levels = c(0, 1), labels = c("Genuine", "Fraud"))
df_2023$Class <- factor(df_2023$Class, levels = c(0, 1), labels = c("Genuine", "Fraud"))

```


```{r}
# Scale the Amount column
#df_2023$Amount <- scale(df_2023$Amount)

# Drop non-feature columns if present (e.g., ID or time)
#df_2023 <- df_2023 %>% select(where(is.numeric))  # keep numeric only

# Convert Class to factor
#df_2023$Class <- factor(df_2023$Class, levels = c(0, 1), labels = c("Genuine", "Fraud"))

# Remove missing values
#df_2023 <- na.omit(df_2023)

```



# ========== EDA ==========

```{r}
# Class Distribution/Class Imbalance
ggplot(df_2013, aes(x = Class)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Class Distribution - 2013 Dataset", x = "Transaction Type", y = "Count")

```

```{r}
# Amount Distribution
ggplot(df_2013, aes(x = Amount)) +
  geom_histogram(bins = 50, fill = "darkgreen") +
  labs(title = "Transaction Amount Distribution (2013)", x = "Scaled Amount", y = "Frequency")

```

```{r}
# PCA Feature Correlation Heatmap
corr_2013 <- cor(df_2013 %>% select(starts_with("V")))
corrplot(corr_2013, method = "color", tl.cex = 0.6, title = "Correlation Heatmap (PCA Features - 2013)")

```

```{r}
# Summary Statistics
summary(df_2013)
```




```{r}
# Class distribution
ggplot(df_2023, aes(x = Class)) +
  geom_bar(fill = "coral") +
  ggtitle("Class Distribution - 2023 Dataset")

```


```{r}
# Transaction Amount distribution
ggplot(df_2023, aes(x = Amount)) +
  geom_histogram(bins = 50, fill = "blue") +
  ggtitle("Transaction Amount Distribution (2023)")
```



```{r}
# Correlation matrix (if PCA-style V1-V28 exist)
pca_cols <- grep("^V[0-9]+$", names(df_2023), value = TRUE)
if (length(pca_cols) > 1) {
  corrplot(cor(df_2023[, pca_cols, with = FALSE]), method = "color", tl.cex = 0.6)
}

```

```{r}
ggplot(df_2023, aes(x = Amount)) +
  geom_histogram(bins = 50, fill = "purple") +
  labs(title = "Transaction Amount Distribution (2023)", x = "Amount", y = "Frequency")

```

```{r}
# Summary
summary(df_2023)
```


# ========== Data Preprocessing ==========

```{r}
ls("package:DMwR2")
```

```{r}
df_balanced <- DMwR2::SMOTE(Class ~ ., data = df_2013, perc.over = 100, perc.under = 200)

```


```{r}
# Install and load DMwR2 if not done
install.packages("DMwR2")
library(DMwR2)

# Apply SMOTE
set.seed(123)
df_balanced <- SMOTE(Class ~ ., data = df_2013, perc.over = 100, perc.under = 200)
table(df_balanced$Class)

```



```{r}
#SMOTE (Synthetic Minority Over-sampling Technique)
set.seed(123)
df_balanced <- SMOTE(Class ~ ., data = df_2013, perc.over = 100, perc.under = 200)
table(df_balanced$Class)

```


```{r}
#Split into Train & Test Sets
set.seed(123)
splitIndex <- createDataPartition(df_balanced$Class, p = 0.8, list = FALSE)
train <- df_balanced[splitIndex, ]
test <- df_balanced[-splitIndex, ]
```

# ========== Model Training and Evaluation ==========

```{r}
# Logistic Regression
log_model <- glm(Class ~ ., data = train, family = binomial)
log_pred <- predict(log_model, test, type = "response")
log_pred_class <- ifelse(log_pred > 0.5, "Fraud", "Genuine")
confusionMatrix(as.factor(log_pred_class), test$Class)

```

```{r}
# Decision Tree
tree_model <- rpart(Class ~ ., data = train, method = "class")
tree_pred <- predict(tree_model, test, type = "class")
confusionMatrix(tree_pred, test$Class)

```

```{r}
# Artificial Neural Network
nn_model <- nnet(Class ~ ., data = train, size = 5, maxit = 200)
nn_pred <- predict(nn_model, test, type = "class")
confusionMatrix(as.factor(nn_pred), test$Class)
```

```{r}
# Gradient Boosting (XGBoost)
train_matrix <- xgb.DMatrix(as.matrix(train[,-ncol(train)]), label = as.numeric(train$Class) - 1)
test_matrix <- xgb.DMatrix(as.matrix(test[,-ncol(test)]))

xgb_model <- xgboost(data = train_matrix, objective = "binary:logistic", nrounds = 100, verbose = 0)
xgb_pred <- predict(xgb_model, test_matrix)
xgb_pred_class <- ifelse(xgb_pred > 0.5, "Fraud", "Genuine")
confusionMatrix(as.factor(xgb_pred_class), test$Class)

```

# ========== Evaluate Models ==========

For all models, calculate:

Accuracy

Precision

Recall

F1-Score

AUPRC

```{r}
# Metrics to Assess
# Example: ROC Curve for Logistic Regression
log_roc <- prediction(log_pred, test$Class)
perf <- performance(log_roc, "tpr", "fpr")
plot(perf, main = "ROC Curve - Logistic Regression")

# AUPRC
pr <- performance(log_roc, "prec", "rec")
plot(pr, main = "Precision-Recall Curve")

```


# ========== Compare Models ==========

Create a table like this:

Model	Accuracy	Precision	Recall	F1-Score	AUPRC
Logistic Regression	XX%	XX%	XX%	XX%	XX
Decision Tree	...	...	...	...	...
ANN	...	...	...	...	...
XGBoost	...	...	...	...	...

```{r}

```


# ========== Interpret Results ==========
📌 Key Points:
Which model performs best overall?

Which model is most interpretable?

How does class imbalance affect results?

Which model is best for real-time deployment?
